---
title: "Jeffrey Chan Final Project - Weather prediction"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tseries)
library(MASS)
library(knitr)
library(lubridate)
library(forecast)
```

# Abstract
In this report, we will analyze a dataset called $\textit{Daily Minimum Temperatures in Melbourne}$[1]. We will use two methods, 1) using 5-day average lowest temperature, and 2) Fourier series to remove the seasonal trend, to fit linear time series models, which can bypass the limitation on the lag number of SARIMA model in R. The results of these two methods are similar, which indicates the validity of both models. In addition, We use both models to forecast the lowest temperature in Melbourne. 

# Introduction
Weather forecast is one of the most common daily tasks in people's life. It helps people to schedule their daily tasks. There are several factors for weather forecasting, one of which is the daily lowest temperature. Although there are meteorological way for forecasting, we are still interested in learning the possibility of predicting using a mathematics-based method. Fortunately, we have learned time series analysis this quarter, which provides us few useful tools to deal with the problem. 

The goal of our project is to use linear time series model to fit and forecast the lowest temperature in Melbourne. We use a dataset $\textit{Daily Minimum Temperatures in Melbourne}$, which contains 3650 (i.e., 10 years) daily temperature data. A seasonal autoregressive integrated moving average (SARIMA) model is often used to analyze this kind of data. However, the lag of seasonal period cannot be greater than 350 in the R software due to the huge amount of memory it will occupy. Therefore, we propose two methods to avoid this constraint. The first method is to use a 5-day average temperature rather than daily data. Using a 5-day average data will result in a seasonal period of $365/5=73$, which is inside the flexible range of R's SARIMA model. Second, we manually remove the seasonal trend and fit the residual using non-seasonal time series model. We will use Fourier series to remove the seasonal trend [2], which is a powerful tool for fitting a curved data and can be easily integrated with ARIMA model. Many libraries have been used in this project: ggplot2, tseries, MASS, knitr, lubridate and forecast. We sincerely thank the authors of these packages. We also thank the R foundation for their contribution to the R community. 
# Experiments

## Load Data

The following code illustrates the procedure to load data. 
```{r}
data <- read.csv("~/temp.csv")
colnames(data) <- c("Date", "Temp")
time <- as_date(data$Date)
target <- data$Temp
```


## Data Visualization 

In order to visually examine the trend and seasonality of the dataset, we first plot the traffic volume against the time.
```{r}
qplot(time, target, ylab="Temperature", xlab="Date")
```

From the plot There is an obvious seasonality in the data, and no obvious trend in the data. 

## Pre-processing

In this section, we will introduce the pre-processing procedure of the data. 

### Deciding the Transformation

Firstly, we would like to see if we need a Box-Cox transformation to pre-process the temperature data. 
```{r}
options(warn=-1)
ts.target = ts(target, frequency = 365)
hist(ts.target, main = "Histogram of Temperature", xlab = "Temperature")
```

The histogram has showed a great range, also has showed a great variation and the distribution is in an approximate bell shape. According to the plot above, we decide not to transform the data.

### Data Aggregation

Since the period used in SARIMA model cannot be greater than 365 in R, we proposed two different methods to fit the data: 1) Using a Fourier series to remove the seasonality in the data, and then fit the non-seasonal part with a normal ARIMA model, 2) Combining five days into one day, i.e. $y_{t} = \frac{(x_{5t} + x_{5t+1}+ x_{5t+2}+ x_{5t+3}+ x_{5t+4})}{5}$, where $x_t$ is the original daily data, and $y_t$ is the newly generated data. 
```{r}
aggregated <- (target[seq(1, length(target), 5)] + 
                      target[seq(2, length(target), 5)] + 
                      target[seq(3, length(target), 5)] + 
                      target[seq(4, length(target), 5)] + 
                      target[seq(5, length(target), 5)] 
                      ) / 5
```

### Spliting Dataset
We split the dataset into two parts, a training set and testing set. The testing set is consisted the last 10% data, and the training set are the remaining.  
```{r}
aggreated_length <- length(aggregated)
aggregated_target <- aggregated[1:(aggreated_length * 0.9)]
aggregated_test <- aggregated[-(1:(aggreated_length * 0.9))]
plot(aggregated_test, type="l", main="Testing Data", ylab="5-day average lowest temperature")
```

## Parameter Identification

In this section, we will find and explain our choice of $p$, $q$, $d$, $P$, $Q$, and $D$. First, we use diff (lag 365/5 = 73) to remove the seasonal trend. 

```{r}
aggregated_target <- ts(aggregated_target, frequency = 73)
diff_aggregated_target <- diff(aggregated_target, 73)
acf(diff_aggregated_target, lag.max=5*73)
pacf(diff_aggregated_target, lag.max=5*73)
```

We can see from the graph above that at seasonal level, PACF seems to tail off, while ACF cuts off at lag 2. It suggest that $P=0$, and $Q=1$. $D=1$, because we difference one time at the seasonal level. Now we need to go back to non-seasonal level. 

```{r}
acf(diff_aggregated_target)
pacf(diff_aggregated_target)
```

The ACF seems to cut off at lag 3, and PACF seems to cut off at lag 2. Therefore, we fit a series of models and pick the best one out of them. Based on this, we pick $p=1$ and $q=2$. 


## Model Fitting
In this section, we show two different methods of model fitting: normal 
### Normal SARIMA Model Fitting
We calculate the value of Akaike’s information criterion (AIC) to evaluate the fitness of a model. Specifically, the formula to calculate AIC is 

$$
\mathrm{AIC}=-2L + 2d,
$$
where $L$ is the likelihood and $d$ is the number of parameters to estimate. Based on our choice of $p$ and $q$, we fit a SARIMA(1,0,2)_(0,0,1)$_73$ model:
```{r}
options(warn=-1)
model <- arima(aggregated_target, order = c(1, 0, 2), seasonal=list(order=c(0,1,1), period=73))
model
```

However, we failed to estimate the standard deviation of seasonal moving average coefficients. Re-fit the model with a smaller moving average order:
```{r}
options(warn=-1)
model <- arima(aggregated_target, order = c(1, 0, 1), seasonal=list(order=c(0,1,1), period=73))
model
```

Although there is an increase in AIC, the results are better than the previous model since the estimated standard deviation of seasonal moving average coefficients is finite.

We compare this model with the model found by auto.arima:
```{r}
auto.model <- auto.arima(aggregated_target)
auto.model
```

It is clear that our model has a smaller AIC compared the model found by auto.arima. Therefore, our choice of model parameters are reasonable. 

The formula of the fitted model is 

$$
(1-0.9410B)(1-B^{73})y_t = (1+0.6951B + 0.1618B^2)(1+0.9937B^{73})z_t 
$$

###  Fourier-series based model
We show in this section that Fourier series can be used to remove the seasonality. 

First, we need to pre-process the daily data. 
```{r}
target_training <- target[time < "1990-01-01"]
target_training <- ts(target_training, frequency = 365)
acf(target_training)
pacf(target_training)
```

It shows a long memory. Difference the data with lag 1:
```{r}
diff_target_training <- diff(target_training)
acf(diff_target_training)
pacf(diff_target_training)
```

ACF cuts off at lag 4, and PACF tails off but peaks at lag 2. So we pick ARMA(3, 2). 

```{r}
fit.fourier <- Arima(target_training, c(3,1,2), xreg=fourier(target_training, K=4))
fit.fourier
```

From the expression, we can see that the seasonal part is modeled by Fourier series, while the non-seasonal part is modeled by an ARIMA(3,1,2). The expression for the non-seasonal part is 

$$
(1+0.1464B-0.2599B^2+0.05B^3)(1-B)y_t=(1-0.307B)z_t
$$ 

## Model Diagnosis

In this section, we will diagnose the model we have just fitted. The first step is to derive the residual of the model.
```{r}
resid <- resid(model)
```

### Residual Visualization

We first examine the trend of the residual visually. 
```{r}
plot(resid)
```

As we can see from the graph, there is no trend in the residual. The variance seems to be uniform all the time. It is also noteworthy that there is a period a time that the residual is very small, since they are basis for prediction. So we remove the residual of the first period from the residual vector. 
```{r}
resid <- resid[-seq(1,73)]
```

Sequentially, we visually examine the normality of the residual:
```{r}
hist(resid)
```

```{r}
qqnorm(resid)
qqline(resid)
```

The Q-Q plot shows that the data are almost on a straight line.

The distribution of the residual is in a bell shape, potentially meaning that the residual is following a normal distribution. The Q_Q plot also supports our judgment on the normality of the residual.  

To be more accurate, we will also use the Sharpio test to numerically examine the normality of the residual. The Sharpio test is a typical test for normality. The test statistics $W$ is calculated as 
$$
W=\frac{(\sum_{i=1}^na_ix_{(i)})^2}{\sum_{i=1}^n(x_i-\bar{x})^2},
$$
where $x_{(i)}$ is the ordered sample values and $a_i$ is a constant generated from the means, variances, and covariances. We will reject normality of the p-value is below 5%.

Testing the normality using Shapiro test: 
```{r}
shapiro.test(resid)
```

The p-value is greater than 0.05, indicating that the residual is following a normal distribution. 

We repeat the same procedure for the Fourier series based model. 
```{r}
resid.fourier <- resid(fit.fourier)
plot(resid.fourier)
resid.fourier <- resid(fit.fourier)
hist(resid.fourier)
```

We do not need to remove the residual of the first period here. The residual is in a bell shape. 

```{r}
qqnorm(resid.fourier)
qqline(resid.fourier)
```

Q-Q plot also shows good normality. 

```{r}
shapiro.test(resid.fourier)
```

However, the residual of the Fourier series-based model cannot pass the normality test. 

### Portmanteau Tests
Next, we will the the independence of residual.

#### Box-Pierce Test
The statistics of the Box-Pierce test is calculated as $$
  Q_W=n\sum_{j=1}^h\hat{\rho}_{\hat{W}}^2(j)\sim\chi^2(h-d)
$$, where $h=\sqrt{n}$, $n$ is the number of data and $d$ is the number of parameters. It is used to examine whether a series is a white noise. We can use the function $Box.test$ to perform Box-Pierce Test:

```{r}
Box.test(resid, lag = as.integer(sqrt(length(resid))) - (1+2+1+1))
```

For the Fourier series-based model:
```{r}
Box.test(resid.fourier, lag = as.integer(sqrt(length(resid.fourier))) - 5)
```

Since both of the p-values are greater than 0.05, we fail to reject the null hypothesis. It suggests that the residuals of the two models are potentially white noise.

#### Ljung-Box Test
The statistics of the Ljung-Box test is calculated as $$
  Q_W=n(n+2)\sum_{j=1}^h\hat{\rho}_{\hat{W}}^2(j)/(n-j)\sim\chi^2(h-d)
$$, where $h=\sqrt{n}$, $n$ is the number of data and $d$ is the number of parameters. It is used to examine whether a series is a white noise. We can use the function $Box.test$ to perform Box-Pierce Test:

```{r}
Box.test(resid, lag = as.integer(sqrt(length(resid))) - 5, type="Ljung-Box")
```

For the Fourier series-based model:
```{r}
Box.test(resid.fourier, lag = as.integer(sqrt(length(resid.fourier))) - 5, type="Ljung-Box")
```

Since both of the p-value is greater than 0.05, we fail to reject the null hypothesis. It suggests that the residuals of the two models are potentially white noise.

#### McLeod Li Test
The statistics of the Ljung-Box test is calculated as 
$$
  Q_W=n(n+2)\sum_{j=1}^h\hat{\rho}_{\hat{W}\hat{W}}^2(j)/(n-j)\sim\chi^2(h-d),
$$
where $h=\sqrt{n}$, $n$ is the number of data and $d$ is the number of parameters. It is used to test the residual for non-linear dependence. 

```{r}
Box.test(resid^2, lag = as.integer(sqrt(length(resid))) - 5, type="Ljung-Box")
```

For the Fourier series-based model:
```{r}
Box.test(resid.fourier^2, lag = as.integer(sqrt(length(resid.fourier))) - 5, type="Ljung-Box")
```

The residual of the SARIMA model has no non-linear dependence but the residual of the Fourier series-based model has. 

From the above diagnosis, we can see that the SARIMA model has a better fitness result, but the Fourier series-based model can handle the daily level data. So each model has its own advantages. 

## Prediction

In this section, we predict the lowest temperature using the two models, i.e. the SARIMA model and the Fourier series-based model. 

### SARIMA Prediction
```{r}
forecast.model <- forecast(model, h = length(aggregated_test))
plot(forecast.model$mean,  ylim=c(0, 20))
lines(ts(aggregated_test, start=c(10,1), frequency=73), col=2)
lines(forecast.model$upper[,2], col=3)
lines(forecast.model$lower[,2], col=3)
```


The red lines are the real 5-day average temperature and the black line is the predicted temperature. The green lines indicate the 95% confidence interval. We can see that the prediction are very close to the real data, and the real data are in the prediction interval. 

### Fourier Prediction
```{r}
target_testing <- ts(target[time >= "1990-01-01"], frequency=365)
forecast.fourier <- forecast(fit.fourier,h = length(target_testing),
        xreg=fourier(target_training, K=4, length(target_testing)))
plot(forecast.fourier$mean,  ylim=c(0, 22))
lines(ts(target_testing, start=c(10,1), frequency=365), col=2)
lines(forecast.fourier$upper[,2], col=3)
lines(forecast.fourier$lower[,2], col=3)
```

The Fourier series-based method would generate very smoothed prediction. The trend of the prediction and the real data are similar; however, the details are different. There are a few days that the real data are outside the 95% confidence interval, showing a inferior performance compared to the SARIMA model. However, given the difficulties of predicting daily temperature due to the high variance, the results are satisfactory. 

# Conclusion

In this report, we successfully modeled the lowest temperatures in Melbourne. In order to bypass the constraints of R that the lag of SARIMA cannot be greater than 350, we have fitted two models (SARIMA model of 5-day average and Fourier series-based model) and showed that these two models can generate good prediction results, i.e. mostly inside the confidence interval of the prediction. 

# Reference
[1] https://www.kaggle.com/paulbrabban/daily-minimum-temperatures-in-melbourne

[2] https://www.rdocumentation.org/packages/forecast/versions/8.13/topics/Arima

# Appendix
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
